---
title: "Literate Analysis with Ocean Water Samples"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

### Introduction

# In this assignment, we will utilize the provided ocean water samples as practice in using the Redfield Ratio. The Redfield Ratio is a significant tool used to measure and diagnose ocean health.  The optimal ratio is known to be 16 nitrogen to 1 phosphate and any fluctuation outside of this ratio serves as an indication of changing conditions due to factors such as polution, climate change, or unusual ocean currents.

# In this assignment, we will utilize the provided ocean water samples as practice in using the Redfield Ratio. The Redfield Ratio is a significant tool used to measure and diagnose ocean health.  The optimal ratio is known to be 16 nitrogen to 1 phosphate and any fluctuation outside of this ratio serves as an indication of changing conditions due to factors such as polution, climate change, or unusual ocean currents.

This Quarto document follows the **Section 5.7 practice exercise** from coreR *Session 05 (Literate Analysis with Quarto)*: building a simple **prose + code** analysis using real seawater chemistry data (`BGchem2008data.csv`).\
Source: <https://learning.nceas.ucsb.edu/2024-10-coreR/session_05.html>

## About the Data

The seawater chemistry dataset used here is `BGchem2008data.csv` (downloaded previously in the course).

-   Date downloaded: August 8, 2025
-   Download source: <https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A35ad7624-b159-4e29-a700-0c0770419941>

You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

## Read in the Data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
bg_chem <- read_csv(here::here("data/raw", "BGchem2008data.csv"))
```

(Optional) Quick peek at the data:

```{r}
# Uncomment if you want more inspection output
 names(bg_chem)
# str(bg_chem)
# head(bg_chem)
# summary(bg_chem)
```

### Analysis

As the “analysis”, we’ll calculate a few simple summary statistics and examine how the nitrogen:phosphate values compare to the **Redfield ratio** concept.

## Calculate Summary Statistics

Compute mean nutrient values (using `na.rm = TRUE` so missing values don’t break the calculation).

```{r}
nitrate <- mean(bg_chem$NO3, na.rm = TRUE)
nitrite <- mean(bg_chem$NO2, na.rm = TRUE)
amm     <- mean(bg_chem$NH4, na.rm = TRUE)
phos    <- mean(bg_chem$P,   na.rm = TRUE)
```

## Calculate Mean Redfield Ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
ratio
```

You can report results inline in Quarto like this:

The Redfield ratio for this dataset is approximately: `r round(ratio)`
# The Redfield Ratio for this dataset is approximately: r round 6.216519


## Plot Redfield Ratio Measurements

Finally, create a simple base R plot that shows the relationship between phosphate and the total nitrogen components.

```{r}
plot(
  bg_chem$P,
  bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4,
  xlab = "Phosphate (P)",
  ylab = "Total nitrogen (NO2 + NO3 + NH4)",
  main = "Ocean water samples: total N vs P"
)
```

### Conclusion

In this mini-analysis, we:

-   Loaded packages in a dedicated setup chunk
-   Read in `BGchem2008data.csv` using a project-root-relative path (`here()`)
-   Calculated simple nutrient means
-   Computed a mean nitrogen:phosphate ratio
-   Created a basic plot relating total nitrogen to phosphate

This is the core idea of **literate analysis**: combine narrative and code in one document so the workflow is easier to understand and reproduce.

# In this project, I begun by installing and loading the appropriate packages such as readr. Then I had to read over the data which initially failed and required me to adjust my code to specify my data/raw folder. After this and the proceeding calculations, I was able to produce a mean nitrogen:phosphate ratio (otherwise known as the Redfield Ratio) of 6.216519. Lastly, I visualized the data by plotting a graph displaying the relationship between phosphate and the total nitrogen components. Very cool!

