---
title: "Analysis of San Bruno Mountain Dataset"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

# In this assignment, I will utilize the provided ocean water samples as practice in using the Redfield Ratio. The Redfield Ratio is a significant tool used to measure and diagnose ocean health.  The optimal ratio is known to be 16 nitrogen to 1 phosphate and any fluctuation outside of this ratio serves as an indication of changing conditions due to factors such as polution, climate change, or unusual ocean currents. *CHANGE THIS TO FIT ASSIGNMENT 2*


## About the data

The coyotebrush datset used here is 'example_data_coyotebrush.csv' (downloaded from prior lab project).

-   Date downloaded: February 23, 2026 
-   Download source: CSV File

We explored the abundance in coyote brush across two different landscapes to identify preferential environmental conditions.


## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(tidyverse)
library(ggplot2)
```

# Read in the data

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
coyotebrush <- read_csv(here::here("data/raw", "example_data_coyotebrush.csv"))
```

(Optional) Quick peek at the data:

```{r}
head(coyotebrush)

summary(coyotebrush)

# Additional viewing functions
# names(coyotebrush)
# str(coyotebrush)
```

# Analysis

## IS THE DATA TIDY?!

Yes! The data does appear to be very tidy, as exemplified in its unique headers, clear numerical inputs, and unit specifications.


# Conclusion

For this assignment, I created a new branch (a02) and a Quarto document (a02.qmd) to evaluate whether my exampled dataset was tidy or not. After importing and inspecting the data, I assessed its organization and variable structure. Based on this review, I determined that the data was in fact tidy as it contained distinct columns, rows, and observational units. Overall, this assignment helped me to practice reproducible research habits that will really help me in with future data analyses.  