---
title: "Analysis of San bruno Mountain Dataset"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

# In this assignment, I will utilize the provided ocean water samples as practice in using the Redfield Ratio. The Redfield Ratio is a significant tool used to measure and diagnose ocean health.  The optimal ratio is known to be 16 nitrogen to 1 phosphate and any fluctuation outside of this ratio serves as an indication of changing conditions due to factors such as polution, climate change, or unusual ocean currents. *CHANGE THIS TO FIT ASSIGNMENT 2*


## About the data

The seawater chemistry dataset used here is `BGchem2008data.csv` (downloaded previously in the course). *CHANGE THIS TO FIT ASSIGNMENT 2*

-   Date downloaded: February 23, 2026 
-   Download source: CSV File

You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

# Read in the data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
bg_chem <- read_csv(here::here("data/raw", "BGchem2008data.csv"))
```

(Optional) Quick peek at the data:

```{r}
# Uncomment if you want more inspection output
 names(bg_chem)
# str(bg_chem)
# head(bg_chem)
# summary(bg_chem)
```

# Analysis

As the “analysis”, we’ll calculate a few simple summary statistics and examine how the nitrogen:phosphate values compare to the **Redfield ratio** concept.

*IS THE DATA TIDY*


## Calculate summary statistics

Compute mean nutrient values (using `na.rm = TRUE` so missing values don’t break the calculation).

```{r}
nitrate <- mean(bg_chem$NO3, na.rm = TRUE)
nitrite <- mean(bg_chem$NO2, na.rm = TRUE)
amm     <- mean(bg_chem$NH4, na.rm = TRUE)
phos    <- mean(bg_chem$P,   na.rm = TRUE)
```
